Estimating Predictive Rate-Distortion via Neural Variational Inference

### 2019-03-05 paper.tex
rewriting

@@

\newcounter{theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{question}[theorem]{Question}
\newtheorem{example}[theorem]{Example}
\newtheorem{defin}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{lemma}[theorem]{Lemma}

@@

@@

@@

Let us assume that an \emph{encoder} $\phi$ is given, expressed in a family of function approximators. % and \emph{decoder} $\psi$ 
The encoder transforms observation sequences $x_{-M\dots -1}$ into a distribution over $Z \in \mathbb{R}^N$: \rljf{Slight mismatch of the math and the prose here: the prose makes it sound like $\phi$ is going to be a function from $x_{-M\dots -1}$ to $Z$ but in the math it returns a probability mass/density function. Maybe write as Z = \phi($x_{-M\dots -1}$), P_\phi(Z|$x_{-M\dots -1}$) = ...}
\begin{equation}
	P(Z|X_{-M\dots M}) := \phi(Z|X_{-M\dots-1}) := \phi(X_{-M\dots-1})(Z)
\end{equation}
From this, one can obtain the conditional distribution over future observations via Bayes' rule, using the Markov condition: \rljf{Show some intermediate steps here}
\begin{equation}\label{eq:cond-fut}
	P(X_{0\dots M-1}|Z) = \frac{\E_{X_{-M \dots M-1}} \phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} \phi(Z|X_{-M \dots -1})}
\end{equation}
However, if $\phi$ is given by a powerful function approximating family such as recurrent neural networks, neither of the two expectations is tractable,  as they require summation over exponentially many sequences. \rljf{Why does this only happen for powerful function approximators? Isn't it the same number of sequences no matter the approximator? Ie doensn't it depend on $M$, not $\phi$?}

@@

\begin{proposition} 
The KL Divergence between $P(Z|X)$ and $q(Z)$, averaged over $X$, upper-bounds the mutual information between $Z$ and the past observations:
\begin{equation}\label{ineq2}
\begin{aligned}
	\mathbb{E}_{X}\left[ D_{KL}(P(Z|X) || q(Z))\right] &=   \mathbb{E}_{X} \mathbb{E}_{Z|X} \log \frac{P_\phi(Z|X_{-M\dots -1})}{q(Z)}  \\
	& \geq  \mathbb{E}_{X} \mathbb{E}_{Z|X} \log \frac{P(Z|X_{-M\dots -1})}{P(Z)}  \\
	& = I[X_{-M\dots 0},Z].
\end{aligned}
\end{equation}
Equality is attained if and only if $q(Z)$ is equal to the true marginal $P(Z) = \E_{X_{-M \dots -1}} \phi(Z|X_{-M \dots -1})$.
\end{proposition}

%These inequalities are tight if $\psi$ is the Bayes optimal decoder ($\psi_z(y) = \sum_{x} \phi(z|x) P(x, y)$), and $q$ is the marginal of $Z$.
% N(\mu(x), \sigma^2(x))
%Note that, for finite $M$, the predictive rate-distortion objective is equivalent to
Therefore, the predictive rate-distortion objective is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q}	\left[-	\mathbb{E}_{z \sim \phi(X_{-M\dots-1})}\left[\log P(X_{0\dots M} | z)\right] + \beta \cdot D_{KL}(\phi(X_{-M\dots-1}) || q)\right],
\end{equation}
where $\phi, \psi, q$ range over all triples of the appropriate types described above.
%As this optimization ranges over all distributions $\phi, \psi, q$ of the appropriate domains, t
To carry out this optimization, we will restrict to a powerful family of parametric families of function approximators, within which we will optimize the objective with gradient descent.
The values obtained for the resulting $\phi, \psi, q$ satisfy the bounds in (\ref{ineq1}-\ref{ineq2}), and---if the family of approximators is sufficiently rich---can come close to turning these into equalities.
%Optimizing them for the objective (\ref{eq:bound}) will result in functions that approximate the optimal decoder and the marginal.
That is, we introduce $\psi$ and $q$ as \emph{variational approximations} \citep{blei-variational-2016} to the true Bayes optimal decoder and marginal. \rljf{Expand last sentence into a paragraph}

@@

NPRD provably provides \emph{upper bounds} on the predictive rate-distortion curve.
This makes it possible to objectively compare the quality of different estimation methods:
Among methods that provide upper bounds, the one that provides the lowest such bound for a given problem is the one giving results closest to the true curve. \rljf{Does the sampling error issue arise here?}

@@

%For OCF, we used $M=1,2...$ until the number of observed sequences was greater than $2 \times 10^4$.

%\paragraph{OCF}

%\paragraph{Neural Method}
%Hyperparameters

%We optimize $q(z|x)$, marginal $s(z)$, and $r(y|z)$.
\paragraph{OCE}
As discussed above, OCE overestimates structure in the process due to overfitting.
To address this, and to enable fair comparison between OCE and our neural method, we follow the evaluation method used in our neural method:
We use held-out data to compute \rljf{Describe these in prose too and what they mean}
$$\frac{1}{N} \sum_{i=1}^N D_{KL}(q(\cdot|x_i) || s(\cdot))$$ %\E_{z \sim q(\cdot|x_i)} \log \frac{q(z|x_i)}{s(z)} $$
and
$$\frac{1}{N} \sum_{i=1}^N \E_{z \sim q(\cdot|x_i)} \log r(y_i|z).$$
Thanks to (\ref{ineq1}-\ref{ineq2}), this provides upper bounds, up to sampling error introduced by finiteness of the held-out data.
Again, this kind of error does not bias the results in either direction, unlike overfitting, which introduces a systematic bias.

In order to be able to apply OCE on held-out data, which will contain sequences that did not occur during estimation, we add a pseudo-sequence $\omega$ and add pseudo-observations $(\omega, x), (x, \omega), (\omega, \omega)$ for all observed sequences $x_{-M...0}, x_{1...M}$ to the matrix of observed counts that serves as the input to the Blahut-Arimoto algorithm.
These pseudo-observations were assigned pseudo-counts $\gamma$ in this matrix of observed counts; we found that a wide range of values ranging from $0.0001$ to $1.0$ yielded essentially the same results.
When evaluating the codebook on held-out data, novel sequences were mapped to $\omega$.
%Given that all unseen futures were mapped to $\omega$, this method may slightly bias towards underestimating entropy of the

%In our experiments, we applied this until the number of sequences was about $20,000$ (agreeing with Marzen and Crutchfield).
%Depending on the topological entropy.

\paragraph{NPRD}
For all experiments, we used $M=15$.

TODO describe

Neural networks have hyperparameters, such as the number of units and the step size used in optimization, that affect the quality of approximation depending on properties of the dataset and the function being approximated.
Given that NPRD provably provides upper bounds, one can identify the best hyperparameters for a given problem by choosing the combination that leads to the lowest estimated upper bounds.

As a computationally more efficient method, we defined a range of plausible hyperparameters based both on experience reported in the literature, and considerations of computational efficiency.
We then randomly sampled, for each of the processes that we experimented on, combinations of $\beta$ and these hyperparameters to run NPRD on.

This method allowed us to both efficiently find good bounds and to estimate the amount of variation in results between the parameter settings.

\begin{tabular}{llll}
	            & Tractable\rljf{??} \\
	Dropout rate & 0.0, 0.1, 0.2 \\
	Embedding Dimension & 50, 100 \\
	LSTM Dimension & 32 \\
	LSTM Layers & 1 \\
	Learning Rate & 0.0001, 0.0005, 0.001 (4 times), 0.002, 0.004 \\
	Input Dropout & 0 \\
	Batch Size & 16, 32, 64 \\
	Flow Length & 1, 2 \\
	Flow Type & DSF, DDSF \\
	Flow Dimension & 32, 64, 128, 512 \\
	Flow Layers & 2 \\
\end{tabular}

\subsection{Analytically Tractable Problems}

We first test NPRD on two processes where the predictive rate-distortion tradeoff is analytically tractable. % by computing the epsilon machine \citep{marzen-predictive-2016}. %, and where prior work has found the epsilon machine.
The \emph{Even Process} \citep{marzen-predictive-2016} is the process of 0/1 IID coin flips, conditioned on all blocks of consecutive ones having even length.
Its complexity and excess entropy are both $\approx 0.63$ nats. % (stationary distribution: 1/3 for odd, 2/3 for even)
It has infinite Markov order, and \cite{marzen-predictive-2016} find that OCE (at $M=5$) performs poorly. %, overestimating  even at $M=5$.
%Their method using the (already known) epsilon machine  works well.
The true predictive rate-distortion curve was computed in \cite{marzen-predictive-2016} using the analytically known epsilon machine.
The \emph{Random Insertion Process} \citep{marzen-predictive-2016} consists of sequences of ones with blocks of at most two zeros interspersed.
The $\epsilon$-machine is given in Figure~\ref{fig:rip-machine} (right).

@@

\paragraph{Recovering Causal States}
Does NPRD lead to inpretable codes $Z$?
To investigate this, we considered the Random Insertion Process. The analytical solution to the predictive rate-distortion problem was computed by \cite{marzen-predictive-2016}.
The process has three causal states.
At $\beta > 0.5$, the optimal solution collapses two of these states in $Z$, while they are separated for $\beta \leq 0.5$.

\rljf{Describe the RIP}

Does NPRD recover this picture?
We applied PCA to samples from $Z$ computed at two different values of $\beta$, $\beta = 0.25$ and $\beta = 0.6$, corresponding to the two points marked in the rate-distortion figure.
The first two principal components are shown in Figure~\ref{fig:latent}.
Samples are colored by the causal states corresponding to the pasts of the trajectories that were encoded into the respective points by NPRD.
On the left, obtained at $\beta=0.6$, two different causal states are collapsed, as expected.
On the right, obtained at $\beta=0.25$, the three causal states are reflected as distinct modes of $Z$.
Note that, at finite $M$, a fraction of pasts is ambiguous between the green and blue causal states; these are colored in black and NPRD maps them into a region between the modes corresponding to these states.

In Figure~\ref{fig:rip-machine}, we record, for each of the three modes where the distribution of the code $Z$ shifts when a symbol is appended.
For comparison, we show the $\epsilon$-machine computed by \cite{marzen-predictive-2016}.
Comparing the two diagrams shows that NPRD effectively recovers the $\epsilon$-machine. \rljf{More detail on this part: it's a bit telegraphic}% even though $Z$ is a multivariate Gaussian variable.

@@

\begin{figure*}
\includegraphics[width=0.45\textwidth]{code/figures/foo_pca_3_machine.png}
\begin{tikzpicture}[shorten >=1pt,node distance=4cm,on grid,auto] 
   \node[state, fill=green] (B)   {$B$}; 
   \node[state, fill=red] (A) [below left=of B] {$A$}; 
   \node[state, fill=blue] (C) [below right=of B] {$C$}; 
    \path[->] 
	(A) edge [bend left=20] node [above] {$\frac{1}{2} | 0$} (B)
	(A) edge [bend left=20] node [below] {$\frac{1}{2} | 1$} (C)
	(B) edge [bend left=20] node [above] {$\frac{1}{2} | 0$} (C)
	(B) edge [bend right=20] node [left] {$\frac{1}{2} | 1$} (C)
	(C) edge [bend left=20] node [left] {$\frac{1}{2} | 1$} (A)
	;
%
%%          edge  node [swap] {1} (B)
%    (C) edge  node  {1} (A)
%          edge [loop above] node {0} ()
%	(A) edge  node [swap] {0} (C) 
%          edge [loop below] node {1} ();
\end{tikzpicture}

@@

%NPRD, in contrasts, encodes the codebook in learned weight matrices that take less than 500MB

%
%This far exceeds what OCE can handle by several orders of magnitude, at least without further optimizations.
%
%Does NPRD 
%
%
%it necessary to increase the number of training iterations to also obtain good estimates of the rate.
%NPRD recovers codebooks with almost zero distortion, though it slightly overestimates the necessary rate.
%NPRD thus finds codebooks that effectively distinguish between millions of different pasts, while requiring orders of magnitude fewer parameters.
%
%The codebook is encoded in weight matrices, takes less than 500MB while training and less when evaluating.
%Directly applying OCE requires creating and storing all $3^{15} \approx 14M$ sequences.

\begin{figure*}
%\includegraphics[width=0.45\textwidth]{code/figures/repeat2-ee-mem.pdf}
\includegraphics[width=0.45\textwidth]{code/figures/repeat3-ee-mem-frontier.pdf}
	\caption{Left: Rate-Distortion for the \textsc{Copy3} process. Green: Curve found by NPRD. Gray: Anaytical curve. \rljf{Red not green? Also y axis should be Memory, not Memories. Also let's label as Memory (nats) throughout (or bits), and Excess Entropy on the x axis throughout.}.}\label{fig:repeat}
\end{figure*}

@@

We consider the problem of estimating rate-distortion for Natural Language. \rljf{Some motivation for why this is cool and interesting, maybe in parallel with the Bentz et al entropy paper, and other information theoretic linguistic complexity papers like Koplenig et al, and Dembowski's work}
%
%As (CITE) have noted, evaluating (X) on finite data will overestimate MI with the future.
%To address this, we estimate (X) on held-out data.
%
%We compare our estimator with OCF on (1) POS-level language modeling, (2) word-level language modeling
%
%Challenges for OCF:
%
%- held-out data will have unobserved sequences.
%We add an out-of-vocabulary pseudo-sequence to the table of observed sequences.
%

@@

Figure~\ref{fig:eng-pos} (right) shows rate as a function of $\beta$.
In Figure~\ref{fig:eng-logbeta}, we show rate and the mutual information with the future, as a function of $\log \beta$.
As $\beta \rightarrow 0$, NPRD (red, $M=15$) continues to discover structure, while OCE (blue, plotted for $M=1,2,3$) exhausts its capacity. \rljf{Remind us what $\beta$ is}

@@

%(The potential number of grammatical 15-word POS sequences is far larger than the corpus -- given the cross-entropies achieved by the LSTMs, the entropy of 15-word English POS sequencs far exceeds 20 nats.)
%The OCE results show that remembering a codebook of longer and longer sequences leads to overfitting.
%Even if computational hurdles were overcome, it seems unlikely that OCE could induce such a large codebook from linguistic corpora, and still maintain generalisability to unseen data.
%How does NPRD manage to get lower distortions?
%This points to an advantage of continuous vector-valued codes, studied extensively in the literature on neural network methods for Natural Language Processing:
%Given the internal structure in code space, it can be expected that, on unseen data, they will construct codes `on the fly' even

@@

Equivalently
\begin{equation}
\lambda = \exp\left(\frac{1}{\alpha^{1/\beta}} M^{1/\beta}\right)
\end{equation}
From this, we can derive expressions for rate $R$ and predictiveness $P$ as follows.
For the solution of Predictive Rate Distortion, we have
\begin{equation}
	\frac{\partial P}{\partial \theta} - \lambda \frac{\partial R}{\partial \theta} = 0,
\end{equation}
where $\theta$ is the codebook defining the code $Z$, and thus
\begin{equation}
\lambda =	\frac{\partial P}{\partial R}.
\end{equation}
%if the rate-predictiveness curve is sufficiently regular at $(P,R)$.

Thus
\begin{equation}\label{eq:derivatives}
%\begin{split}
	\partial P/\partial R = \exp\left(-\frac{1}{\alpha^{1/\beta}} R^{1/\beta}\right). %&\ \ \ \partial R/\partial P = \exp\left(\frac{1}{\alpha^{1/\beta}} R^{1/\beta}\right)
%\end{split}
\end{equation}
Qualitatively, this says that $P$ asymptotes to a finite value, whereas $R$ is unbounded. \rljf{Exactly the opposite of Dembowski's finding}

%
%\begin{equation}
%P = \int \exp\left(-\frac{1}{\alpha^{1/\beta}} R^{1/\beta}\right) dR
%\end{equation}
%
%\begin{equation}
%R = \int \exp\left(-\frac{1}{\alpha^{1/\beta}} R^{1/\beta}\right) dP
%\end{equation}
%

%Pretending that $t$ is continuous, we write
%\begin{equation}
%	E = \int_1^{M/c} t f(t)
%\end{equation}
%and thus
%\begin{equation}
%	t f(M/c) = \partial E/\partial M  = \exp\left(-\frac{1}{\alpha^{1/\beta}} M^{1/\beta}\right)
%\end{equation}
%and thus
%\begin{equation}
%	f(t) = \frac{1}{t} \exp\left(-\left(\frac{c}{\alpha}\right)^{1/\beta} t^{1/\beta}\right)
%\end{equation}
%We get
%\begin{equation}
%E = \int \exp\left(-\frac{1}{\alpha^{1/\beta}} M^{1/\beta}\right) dM
%\end{equation}
%
%

%
%
%Why would language show this behavior?
%We will describe a proposal.
%%We will describe a proposal that leads to quantitative predictions.
%As a simple model of prediction, we consider a code $Z^{(T)}$ that stores all relevant predictive information from the past $T$ words (where $T$ depends on $\lambda$), and nothing from the words further in the past.
%The predictiveness of this code is
%\begin{equation}
%	I[X_{-T...-1}, \overrightarrow{X}]
%\end{equation}
%
%Let us make the assumption that storing the relevant information from each past word requires maintaining an amount of information about each word that is bounded away from zero:
%$$I[X_{-t}, Z^{(T)}|X_{-t+1}\dots X_{-1}] = c + g(t)$$
%for each $t = 1, ..., T$, where $c > 0$ and $g(t) \geq 0$.
%We will ignore $g(t)$ in the following.
%Then storing relevant information about the last $T$ words will lead to a rate of $\approx cT$, and predictiveness of 
%\begin{equation}
%	E = \sum_{t=1}^T t f(t)
%\end{equation}
%
%
%Let $f(t)$ be the incremental contributions of past words to predicting the next word:
%$$f(t) := I[X_t, X_0|X_1\dots X_{t-1}]$$
%(Since $H[X_0]$ is finite, $f(t)$ must go to zero fast enough to be summable, even in the idealized setting of infinite contexts.)

>>>>

\usepackage[T1]{fontenc}

@@

\newcounter{theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{question}[theorem]{Question}
\newtheorem{example}[theorem]{Example}
\newtheorem{defin}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{thm}[theorem]{Theorem}

@@

\usepackage{mathtools}

\DeclarePairedDelimiterX{\infdivx}[2]{(}{)}{%
  #1\;\delimsize\|\;#2%
}
\newcommand{\infdiv}{D\infdivx}
\DeclarePairedDelimiter{\norm}{\lVert}{\rVert}

@@

GENERAL INTRO/FRAMING

@@

Let us assume that an \emph{encoder} $\phi$ is given, expressed in a family of function approximators. % and \emph{decoder} $\psi$ 
The encoder stochastically transforms observation sequences $x_{-M\dots -1}$ into a random code $Z \in \mathbb{R}^N$: \rljf{Slight mismatch of the math and the prose here: the prose makes it sound like $\phi$ is going to be a function from $x_{-M\dots -1}$ to $Z$ but in the math it returns a probability mass/density function. Maybe write as Z = \phi($x_{-M\dots -1}$), P_\phi(Z|$x_{-M\dots -1}$) = ...}
\begin{equation}
%	P(Z|X_{-M\dots M}) := \phi(Z|X_{-M\dots-1}) := \phi(X_{-M\dots-1})(Z)
Z = \phi(x_{-M\dots -1})
\end{equation}
distributed according to $Z \sim P_\phi(Z|x_{-M\dots -1})$.
From this, one can obtain the conditional distribution over future observations via Bayes' rule: %, using the Markov condition: \rljf{Show some intermediate steps here}
%\begin{equation}\label{eq:cond-fut}
%	P(X_{0\dots M-1}|Z) = \frac{\E_{X_{-M \dots M-1}} P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})}
%\end{equation}
%
\begin{equation}
	\label{eq:cond-fut}
	\begin{split}
	P(X_{0\dots M-1}|Z) &= \frac{p(X_{0\dots M-1}, Z)}{p(Z)} \\
	&= \frac{\E_{X_{-M \dots -1}} P_\phi(X_{0\dots M-1}, Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})} \\
	&=^{(*)} \frac{\E_{X_{-M \dots -1}}   P(X_{0\dots M-1}|X_{-M\dots-1})     P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})} \\
	&= \frac{\E_{X_{-M \dots M-1}} P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})}
	\end{split}
\end{equation}
where $(*)$ uses the Markov condition.
% if $\phi$ is given by a powerful function approximating family such as recurrent neural networks, 
However, neither of the two expectations is tractable,  as they require summation over exponentially many sequences.
If $\phi$ is given by a powerful function approximating family such as recurrent neural networks, algorithms (e.g., dynamic programming) to compute this sum efficiently are not available.

@@

\begin{proposition} 
The KL Divergence between $P(Z|X)$ and $q(Z)$, averaged over $X$, upper-bounds the mutual information between $Z$ and the past observations:
\begin{equation}\label{ineq2}
\begin{aligned}
	\mathbb{E}_{X}\left[ \operatorname{D_{KL}}(\infdiv{P(Z|X)}{q(Z)})\right] &=   \mathbb{E}_{X} \mathbb{E}_{Z|X} \log \frac{P_\phi(Z|X_{-M\dots -1})}{q(Z)}  \\
	& \geq  \mathbb{E}_{X} \mathbb{E}_{Z|X} \log \frac{P(Z|X_{-M\dots -1})}{P(Z)}  \\
	& = I[X_{-M\dots 0},Z].
\end{aligned}
\end{equation}
Equality is attained if and only if $q(Z)$ is equal to the true marginal $P(Z) = \E_{X_{-M \dots -1}} \phi(Z|X_{-M \dots -1})$.
\end{proposition}

Therefore, the predictive rate-distortion objective is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q}	\left[-	\mathbb{E}_{z \sim \phi(X_{-M\dots-1})}\left[\log P(X_{0\dots M} | z)\right] + \beta \cdot \operatorname{D_{KL}}(\infdiv{\phi(X_{-M\dots-1})}{q})\right],
\end{equation}
where $\phi, \psi, q$ range over all triples of the appropriate types described above.
To carry out this optimization, we will restrict to a powerful family of parametric families of function approximators, within which we will optimize the objective with gradient descent.
The values obtained for the resulting $\phi, \psi, q$ satisfy the bounds in (\ref{ineq1}-\ref{ineq2}), and---if the family of approximators is sufficiently rich---can come close to turning these into equalities.

That is, we introduce $\psi$ and $q$ as \emph{variational approximations} \citep{blei-variational-2016} to the true Bayes optimal decoder and marginal. \rljf{Expand last sentence into a paragraph}

@@

NPRD provably provides \emph{upper bounds} on the predictive rate-distortion curve, up to sampling error.
This makes it possible to objectively compare the quality of different estimation methods:
Among methods that provide upper bounds, the one that provides the lowest such bound for a given problem is the one giving results closest to the true curve (up to sampling error).

@@

\paragraph{OCE}
As discussed above, OCE overestimates structure in the process due to overfitting.
To address this, and to enable fair comparison between OCE and our neural method, we follow the evaluation method used for NPRD, evaluating rate and distortion on held-out data:
%We use held-out data to compute empirical estimates of quantities anlogous to those computed by NPRD.
\rljf{Describe these in prose too and what they mean}
We estimate rate using the variational bound that we stated in Proposition~\ref{ineq2}:
\begin{equation}
\frac{1}{N} \sum_{i=1}^N \operatorname{D_{KL}}(\infdiv{q(\cdot|x_i)}{s(\cdot)})
\end{equation}
%\E_{z \sim q(\cdot|x_i)} \log \frac{q(z|x_i)}{s(z)} $$
We estimate the prediction loss on the future observations as the empirical cross-entropy, i.e., the variational bound stated in Proposition~\ref{ineq1}:
\begin{equation}
\frac{1}{N} \sum_{i=1}^N \E_{z \sim q(\cdot|x_i)} \log r(y_i|z).
\end{equation}
Thanks to (\ref{ineq1}-\ref{ineq2}), these quantities provide upper bounds, up to sampling error introduced by finiteness of the held-out data.
Again, this kind of error does not bias the results in either direction, unlike overfitting, which introduces a systematic bias.

In order to be able to apply OCE on held-out data, which will contain sequences that did not occur during estimation, we add a pseudo-sequence $\omega$ and add pseudo-observations $(\omega, x), (x, \omega), (\omega, \omega)$ for all observed sequences $x_{-M...0}, x_{1...M}$ to the matrix of observed counts that serves as the input to the Blahut-Arimoto algorithm.
These pseudo-observations were assigned pseudo-counts $\gamma$ in this matrix of observed counts; we found that a wide range of values ranging from $0.0001$ to $1.0$ yielded essentially the same results.
When evaluating the codebook on held-out data, novel sequences were mapped to $\omega$.

\paragraph{NPRD}
For all experiments, we used $M=15$.
Neural networks have hyperparameters, such as the number of units and the step size used in optimization, that affect the quality of approximation depending on properties of the dataset and the function being approximated.
Given that NPRD provably provides upper bounds, one can identify the best hyperparameters for a given problem by choosing the combination that leads to the lowest estimated upper bounds.

As a computationally more efficient method, we defined a range of plausible hyperparameters based both on experience reported in the literature, and considerations of computational efficiency.
These parameters are shown in Table~\ref{tab:nprd-hyperparameters}.
We then randomly sampled, for each of the processes that we experimented on, combinations of $\beta$ and these hyperparameters to run NPRD on.

The first group of parameters are the dimensions of the input embedings, the recurrent LSTM states.
%We only use
The second group of parameters are regularization parameters.
We apply dropout~\citep{srivastava-dropout:-2014} to the input and output layers.
The third group of parameters are related to the optimization procedure.
We optimize all parameters using Adam~\citep{kingma-adam:-2014}, a variant of stochastic gradient descent~\citep{robbins1951stochastic} commonly used for optimizing neural networks.
We implemented the model using PyTorch \citep{paszke2017automatic}.
Neural Autoregressive Flows, used to approximate the marginal $q$, also have a set of hyperparameters \citep{huang-neural-2018}: The length of the flow, the type (DSF / Deep Sigmoid Flow or  DDSF / Deep Dense Sigmoid Flow), the dimension of the flow, and the number of layers in the flow.

%This method allowed us to both efficiently find good bounds and to estimate the amount of variation in results between the parameter settings.

\begin{table}
\begin{tabular}{l||lll}
	            & Section~\ref{sec:tractable} \\ \hline\hline
	Embedding Dimension & 50, 100 \\
	LSTM Dimension & 32 \\ \hline
%	LSTM Layers & 1 \\ \hline
	Dropout rate & 0.0, 0.1, 0.2 \\
	Input Dropout & 0 \\ \hline
	Adam Learning Rate & 0.0001, 0.0005, 0.001 (4 times), 0.002, 0.004 \\
	Batch Size & 16, 32, 64 \\ \hline
	Flow Length & 1, 2 \\
	Flow Type & DSF, DDSF \\
	Flow Dimension & 32, 64, 128, 512 \\
	Flow Layers & 2 \\
\end{tabular}
	\caption{NPRD Hyperparameters}\label{tab:nprd-hyperparameters}
\end{table}

\subsection{Analytically Tractable Problems}\label{sec:tractable}

We first test NPRD on two processes where the predictive rate-distortion tradeoff is analytically tractable. % by computing the epsilon machine \citep{marzen-predictive-2016}. %, and where prior work has found the epsilon machine.
The \emph{Even Process} \citep{marzen-predictive-2016} is the process of 0/1 IID coin flips, conditioned on all blocks of consecutive ones having even length.
Its complexity and excess entropy are both $\approx 0.63$ nats. % (stationary distribution: 1/3 for odd, 2/3 for even)
It has infinite Markov order, and \cite{marzen-predictive-2016} find that OCE (at $M=5$) performs poorly. %, overestimating  even at $M=5$.
%Their method using the (already known) epsilon machine  works well.
The true predictive rate-distortion curve was computed in \cite{marzen-predictive-2016} using the analytically known epsilon machine.
The \emph{Random Insertion Process} \citep{marzen-predictive-2016} consists of sequences of uniform coin flips $X_t \in \{0,1\}$, subject to the constraint that, if $X_{t-2}$ was a 0, then $X_{t}$ has to be a 1.

@@

\paragraph{Recovering Causal States}
Does NPRD lead to inpretable codes $Z$?
To answer this, we further investigated the NPRD approximation to the Random Insertion Process (RIP), obtained in the previous paragraph.
The $\epsilon$-machine was computed by  \cite{marzen-predictive-2016} and is given in Figure~\ref{fig:rip-machine} (right).
The process has three caual states:
A represents those pasts where the future starts with $1^{2k}0$ ($k = 0, 1, 2, \dots$) -- these are the pasts ending in either $001$ or $10111^m$ ($m= 0, 1, 2, \dots$). % but cannot go in with 10
B represents those pasts ending in 10 -- the future has to start with 01 or 11.
C represents those pasts ending in either 00 or 01 -- the future has to start with $1^{2k+1}0$ ($k = 0, 1, 2, \dots$).
%These states represent the 
%As itThis $\epsilon$-machine is the smallest unifilar

The analytical solution to the predictive rate-distortion problem was computed by \cite{marzen-predictive-2016}.
At $\beta > 0.5$, the optimal solution collapses $A$ and $B$ into a single codeword, while all three states are mapped to separate codewords for $\beta \leq 0.5$.

Does NPRD recover this picture?
We applied PCA to samples from $Z$ computed at two different values of $\beta$, $\beta = 0.25$ and $\beta = 0.6$, corresponding to the two points marked in the rate-distortion figure.
The first two principal components are shown in Figure~\ref{fig:latent}.
Samples are colored by the causal states corresponding to the pasts of the trajectories that were encoded into the respective points by NPRD.
On the left, obtained at $\beta=0.6$, the states A and B are collapsed, as expected.
On the right, obtained at $\beta=0.25$, the three causal states are reflected as distinct modes of $Z$.
Note that, at finite $M$, a fraction of pasts is ambiguous between the green and blue causal states; these are colored in black and NPRD maps them into a region between the modes corresponding to these states.

In Figure~\ref{fig:rip-machine}, we record, for each of the three modes to which cluster the distribution of the code $Z$ shifts when a symbol is appended.
We restrict to those strings that have nonzero probability for RIP (no code will ever be needed for other strings).
For comparison, we show the $\epsilon$-machine computed by \cite{marzen-predictive-2016}.
Comparing the two diagrams shows that NPRD effectively recovers the $\epsilon$-machine:
The three causal states are represented by the three different modes of $Z$, and the effect of appending a symbol also mirrors the state transitions of the $\epsilon$-machine.
%\rljf{More detail on this part: it's a bit telegraphic}% even though $Z$ is a multivariate Gaussian variable.

@@

\begin{figure*}
\includegraphics[width=0.45\textwidth]{code/figures/foo_pca_3_machine.png}
\begin{tikzpicture}[shorten >=1pt,node distance=4cm,on grid,auto] 
   \node[state, fill=green] (B)   {$B$}; 
   \node[state, fill=red] (A) [below left=of B] {$A$}; 
   \node[state, fill=blue] (C) [below right=of B] {$C$}; 
    \path[->] 
	(A) edge [bend left=20] node [above] {$\frac{1}{2} | 0$} (B)
	(A) edge [bend left=20] node [below] {$\frac{1}{2} | 1$} (C)
	(B) edge [bend left=20] node [above] {$\frac{1}{2} | 0$} (C)
	(B) edge [bend right=20] node [left] {$\frac{1}{2} | 1$} (C)
	(C) edge [bend left=20] node [left] {$1 | 1$} (A)
	;
%
%%          edge  node [swap] {1} (B)
%    (C) edge  node  {1} (A)
%          edge [loop above] node {0} ()
%	(A) edge  node [swap] {0} (C) 
%          edge [loop below] node {1} ();
\end{tikzpicture}

@@

\begin{figure*}
%\includegraphics[width=0.45\textwidth]{code/figures/repeat2-ee-mem.pdf}
\includegraphics[width=0.45\textwidth]{code/figures/repeat3-ee-mem-frontier.pdf}
	\caption{Rate-Distortion for the \textsc{Copy3} process. We show NPRD samples, and the resulting upper bound in red. The gray line represents the anaytical curve.}\label{fig:repeat}
\end{figure*}
%  \rljf{Red not green? Also y axis should be Memory, not Memories. Also let's label as Memory (nats) throughout (or bits), and Excess Entropy on the x axis throughout.}

@@

We consider the problem of estimating rate-distortion for Natural Language.

TODO Motivation

UID

%Fenk, A.; Fenk, G. Konstanz im Kurzzeitgedächtnis-Konstanz im sprachlichen Informationsfluß. Z. Exp. Angew. Pshychol. 1980, 27, 400–414.
%Fenk-Oczlon, G. Familiarity, information flow, and linguistic form. In Frequency and the Emergence of Linguistic Structure; Bybee, J.L., Hopper, P.J., Eds.; John Benjamins: Amsterdam, The Netherlands, 2001; pp. 431–448.
%Ferrer-i Cancho, R.; D ̨ebowski, Ł.; del Prado Martín, F.M. Constant conditional entropy and related hypotheses. J. Stat. Mech. Theory Exp. 2013, L07001, doi:10.1088/1742-5468/2013/07/L07001.

People have studied information density as a model of cost in human language processing.

But also memory cost is relevant

\cite{bentz2017entropy}

\cite{koplenig2017statistical}

\cite{dkebowski2018natural}

\cite{takahira2016entropy}

\rljf{Some motivation for why this is cool and interesting, maybe in parallel with the Bentz et al entropy paper, and other information theoretic linguistic complexity papers like Koplenig et al, and Dembowski's work}

@@

Figure~\ref{fig:eng-pos} (right) shows rate as a function of $\beta$.
In Figure~\ref{fig:eng-logbeta}, we show rate and the mutual information with the future, as a function of $\log \frac{1}{\beta}$.
Recall that $\beta$ is the tradeoff-parameter from the objective function~\ref{eq:ib}.
As $\beta \rightarrow 0$, NPRD (red, $M=15$) continues to discover structure, while OCE (blue, plotted for $M=1,2,3$) exhausts its capacity.

@@

@@

Equivalently
\begin{equation}
\lambda = \exp\left(\frac{1}{\alpha^{1/\beta}} M^{1/\beta}\right)
\end{equation}
From this, we can derive expressions for rate $R$ and predictiveness $P$ as follows.
For the solution of Predictive Rate Distortion, we have
\begin{equation}
	\frac{\partial P}{\partial \theta} - \lambda \frac{\partial R}{\partial \theta} = 0,
\end{equation}
where $\theta$ is the codebook defining the code $Z$, and thus
\begin{equation}
\lambda =	\frac{\partial P}{\partial R}.
\end{equation}

Thus
\begin{equation}\label{eq:derivatives}
	\partial P/\partial R = \exp\left(-\frac{1}{\alpha^{1/\beta}} R^{1/\beta}\right). %&\ \ \ \partial R/\partial P = \exp\left(\frac{1}{\alpha^{1/\beta}} R^{1/\beta}\right)
\end{equation}
Qualitatively, this says that $P$ asymptotes to a finite value, whereas $R$ is unbounded. \rljf{Exactly the opposite of Dembowski's finding}
